---
title: "Exploratory Data Analysis, Linear Regression, and Prediction Modeling of Student Performance"
author: "Travis Gubbe"
date: "September 25, 2022"
output: rmarkdown::github_document
---

In this R Markdown session, I will use various methods of exploratory data analysis to examine the characteristics of the "Students Performance in Exams" dataset. In addition, linear regression to view relationships within the data. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(car)
library(ggplot2) #Used for visual plots.
library(pander) #Used to create table.
```

## About the Dataset

The dataset used can be found on the Kaggle website, an online platform for data scientists containing free datasets and code collaboration. Below is the link for the dataset:
<https://www.kaggle.com/datasets/spscientist/students-performance-in-exams>.

The data contains eight variables used to explore the effect of various factors on test scores. The variables are: 

* Gender - male or female
* Race/Ethnicity - split into 5 Groups (A, B, C, D, E)
* Parent Education Level - a student's parent's highest level of education
* Lunch - whether a student is in the standard or free/reduced lunch program
* Test Prep Course - whether or not a student completed the test preparation course
* Math Score
* Reading Score
* Writing Score

**Note: this is a fictional dataset used strictly for to demonstrate beginner data analysis skills. The results are not official and should not be used to conclude actual relationships between the variables listed and education**

First, the dataset is loaded into R and saved as "student".

```{r dataset}
student <- read.csv("~/R datasets/StudentsPerformance.csv")
```

## Viewing and Cleaning the Data

Now that the dataset is loaded into R, the next step is to view the data and see if it's clean for analysis. 

```{r data_summary}
#Inspect the data frame.
head(student)
#View the column names.
colnames(student)
#View summary of the data frame.
summary(student)
#View data types in the data frame.
str(student)
```

I don't like the "." in the names, electing to change the "." to a "_" for easier reading. Also, the name of the "race_ethnicity" variable is shortened to "ethnicity". 

```{r rename_columns}
#Rename the columns in the student data frame.
student <- student%>%rename(parental_education = parental.level.of.education, math_score = math.score, writing_score = writing.score, reading_score = reading.score, ethnicity = race.ethnicity, test_prep_course = test.preparation.course)
#View the updated column names in the student data frame.
colnames(student)
```

Next, I want to check if there are any missing values in the dataset.

```{r missing_values}
#Print the total number of missing values in the data frame.
sum(is.na(student))
```

Now that I know there aren't any missing values, next I check to see if there are any duplicates in the dataset.

```{r duplicates}
#Create a variable storing the amount of duplicates in the data frame.
duplicates <- student%>%duplicated()
#Displays how many duplicates are present in a table. If a value is not a duplicate, it is placed in 'FALSE'. If the value is a duplicate, it is placed in 'TRUE'.
duplicates_count <- duplicates%>%table()
duplicates_count
```

So far, there are no missing values or duplicates in the data. Next, I want to view the distribution of the data for each variable.

First, a count and frequency table is created to see the distribution of the data.

```{r data_count}
#The total number of males and females in the data frame.
count_gender <- student%>%count(gender)
count_gender
#Create a frequency table to show the percentage of each gender in the data frame.
freq_gender <- table(student$gender)/length(student$gender)
freq_gender
#The total number of each ethnicity group in the data frame.
count_ethnicity <- student%>%count(ethnicity)
count_ethnicity
#Creates a frequency table to show the percentage of each ethnicity group in the data frame.
freq_ethnicity <- table(student$ethnicity)/length(student$ethnicity)
freq_ethnicity
#The total number of each parental highest education group in the data frame.
count_parental_education <- student%>%count(parental_education)
count_parental_education
#Creates a frequency table to show the percentage of each parental highest education group in the data frame.
freq_parental <- table(student$parental_education)/length(student$parental_education)
freq_parental
#The total number of students who took the test prep course.
count_test_prep <- student%>%count(test_prep_course)
count_test_prep
#Creates a frequency table to show the percentage of each parental highest education group in the data frame.
freq_test <- table(student$test_prep_course)/length(student$test_prep_course)
freq_test
#The total number of students in each lunch group.
count_lunch <- student%>%count(lunch)
count_lunch
#Creates a frequenct table to show the percentage of each lunch group.
freq_lunch <- table(student$lunch)/length(student$lunch)
freq_lunch
```

Boxplots can also be used to view the distribution of the data in each test score. The boxplot will show the interquartile range (IQR), mean, and outliers in the data.

```{r boxplots}
boxplot(student$reading_score)
boxplot(student$writing_score)
boxplot(student$math_score)
```

From the boxplots, there appear to be multiple outliers in each test score variable. The one I'm most concerned with is the "0" in the math test section. 

I want to find out which student scored a 0 on the math exam.

```{r outlier_location}
which.min(student$math_score)
```

Row 60 (student #60) contains the student who scored a "0" on the math test. In addition, I want to view the values of student #60 to see if all of their exam results and to determine if they are an outlier in the data frame:

```{r view_outlier}
#View student 60 to see their overall test results and determine if they're an outlier.
print(student[60, ])

```

Student #60 has very low scores in each test. I can check the range of each test to see if student #60 has the lowest scores in each test:

```{r max_min}
range(student$reading_score)

range(student$writing_score)

range(student$math_score)
```

Student #60 has the lowest test score in each section. To me, student #60 is an outlier and can be removed from the data. 

```{r remove_outlier}
#Remove the 0 test score from the data set. Will also remove the student completely from the data frame.
student<- subset(student, math_score != 0)
#Counts the amount of rows in the data to see if the outlier was properly removed.
count(student)
#Count the amount in each gender. Compared to the original count, the female count lowered by 1.
count_gender <- student%>%count(gender)
count_gender
```

### Viewing the Distribution of the Data Among Scores

With the data inspected and cleaned, histograms are created to see the distribution of the variables. First, the distribution of the test scores are viewed using ggplot2.

```{r histograms}
#Histogram displaying the distribution of the Reading Scores for the data frame
ggplot(data = student, aes(reading_score))+geom_histogram(aes(y = ..density..), color = "darkblue", fill = "lightblue", binwidth = 5)+labs(title = "Distribution of Reading Scores", x = "Reading Score", y = "Total Count")+geom_density(alpha = 0.3, color = "skyblue", fill = "skyblue")
#Histogram displaying the distribution of the Writing Scores for the data frame
ggplot(data = student, aes(writing_score))+geom_histogram(aes(y = ..density..), color = "darkgreen", fill = "lightgreen", binwidth = 5)+labs(title = "Distribution of Writing Scores", x = "Writing Score", y = "Total Count")+geom_density(alpha = 0.2, color = "darkgreen", fill = "lightgreen")
#Histogram displaying the distribution of the Math Scores for the data frame
ggplot(data = student, aes(math_score))+geom_histogram(aes(y = ..density..), color = "darkred", fill = "red", binwidth = 5)+labs(title = "Distribution of Math Scores", x = "Math Score", y = "Total Count")+geom_density(alpha = 0.2, color = "black", fill = "red")
```

Next, the distribution of gender in each test is viewed using ggplot2.

```{r hist_gender}
#Histogram displaying distribution of reading scores between gender
ggplot(student, aes(x = reading_score))+geom_histogram(aes(color = gender, fill = gender), position = "identity", binwidth = 5, alpha = 0.4)+scale_color_manual(values = c("#00AFBB", "#E7B800"))+scale_fill_manual(values = c("#00AFBB", "#E7B800"))+labs(title = "Distribution of Reading Scores Between Gender", x = "Reading Score", y = "Total Count")

#Histogram displaying distribution of writing scores between gender
ggplot(student, aes(x = writing_score))+geom_histogram(aes(color = gender, fill = gender), position = "identity", binwidth = 5, alpha = 0.4)+scale_color_manual(values = c("#00AFBB", "#E7B800"))+scale_fill_manual(values = c("#00AFBB", "#E7B800"))+labs(title = "Distribution of Writing Scores Between Gender", x = "Writing Score", y = "Total Count")

#Histogram displaying distribution of math scores between gender
ggplot(student, aes(x = math_score))+geom_histogram(aes(color = gender, fill = gender), position = "identity", binwidth = 5, alpha = 0.4)+scale_color_manual(values = c("#00AFBB", "#E7B800"))+scale_fill_manual(values = c("#00AFBB", "#E7B800"))+labs(title = "Distribution of Math Scores Between Gender", x = "Math Score", y = "Total Count")
```

Overall, the data appears to be normally distributed. 

## Boxplots of Each Variable

Next, boxplots are created to compare the test scores to the categorical variables. A boxplot comparison is done for gender, race/ethnicity, parental education level, lunch program, and test preparation.

```{r box_gender}
ggplot(student, aes(gender, reading_score, fill = gender, color = gender))+geom_boxplot()+labs(title = "Reading Score for Each Gender", x = "Gender", y = "Reading Score")+scale_fill_manual(values = c("purple", "lightblue"))+scale_color_manual(values = c("black", "gray5"))

ggplot(student, aes(gender, writing_score, fill = gender, color = gender))+geom_boxplot()+labs(title = "Writing Score for Each Gender", x = "Gender", y = "Reading Score")+scale_fill_manual(values = c("purple", "lightblue"))+scale_color_manual(values = c("black", "gray5"))

ggplot(student, aes(gender, math_score, fill = gender, color = gender))+geom_boxplot()+labs(title = "Math Score for Each Gender", x = "Gender", y = "Math Score")+scale_fill_manual(values = c("purple", "lightblue"))+scale_color_manual(values = c("black", "gray5"))
```

```{r box_ethnicity}
ggplot(student, aes(ethnicity, reading_score, fill = ethnicity, color = ethnicity))+geom_boxplot()+labs(title = "Reading Score for Each Ethnicity Group", x = "Ethnicity", y = "Reading Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1"))

ggplot(student, aes(ethnicity, writing_score, fill = ethnicity, color = ethnicity))+geom_boxplot()+labs(title = "Writing Score for Each Ethnicity Group", x = "Ethnicity", y = "Writing Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1"))

ggplot(student, aes(ethnicity, math_score, fill = ethnicity, color = ethnicity))+geom_boxplot()+labs(title = "Math Score for Each Ethnicity Group", x = "Ethnicity", y = "Math Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1"))
```

```{r box_education}
ggplot(student, aes(parental_education, reading_score, fill = parental_education, color = parental_education))+geom_boxplot()+labs(title = "Reading Score for Each Parental Education Group", x = "Parental Education Level", y = "Reading Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))

ggplot(student, aes(parental_education, writing_score, fill = parental_education, color = parental_education))+geom_boxplot()+labs(title = "Writing Score for Each Parental Education Group", x = "Parental Education Level", y = "Writing Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))

ggplot(student, aes(parental_education, math_score, fill = parental_education, color = parental_education))+geom_boxplot()+labs(title = "Math Score for Each Parental Education Group", x = "Parental Education Level", y = "Math Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))
```

```{r box_lunch}
ggplot(student, aes(lunch, reading_score, fill = lunch, color = lunch))+geom_boxplot()+labs(title = "Reading Score for Each Parental Education Group", x = "Lunch Program", y = "Reading Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))

ggplot(student, aes(lunch, writing_score, fill = lunch, color = lunch))+geom_boxplot()+labs(title = "Writing Score for Each Lunch Program Group", x = "Lunch Program", y = "Writing Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))

ggplot(student, aes(lunch, math_score, fill = lunch, color = lunch))+geom_boxplot()+labs(title = "Math Score for Each Parental Education Group", x = "Lunch Program", y = "Math Score")+scale_fill_manual(values = c("blue", "red", "yellow", "orange", "purple", "darkgoldenrod1"))+scale_color_manual(values = c("hotpink", "purple3", "black", "royalblue3", "violetred1", "burlywood4"))
```

```{r box_test_prep}
ggplot(student, aes(test_prep_course, reading_score, fill = test_prep_course, color = test_prep_course))+geom_boxplot()+labs(title = "Reading Score for Each Test Preparation Group", x = "Test Preparation - Yes or No", y = "Reading Score")+scale_fill_manual(values = c("blue", "red"))+scale_color_manual(values = c("hotpink", "royalblue3"))

ggplot(student, aes(test_prep_course, writing_score, fill = test_prep_course, color = test_prep_course))+geom_boxplot()+labs(title = "Writing Score for Each Test Preparation Group", x = "Test Preparation - Yes or No", y = "Writing Score")+scale_fill_manual(values = c("blue", "red"))+scale_color_manual(values = c("hotpink", "royalblue3"))

ggplot(student, aes(test_prep_course, math_score, fill = test_prep_course, color = test_prep_course))+geom_boxplot()+labs(title = "Math Score for Each Test Preparation Group", x = "Test Preparation - Yes or No", y = "Math Score")+scale_fill_manual(values = c("blue", "red"))+scale_color_manual(values = c("hotpink", "royalblue3"))
```

In each boxplot, there appears to be a difference in means between each group as related to reading, writing, and math test scores. 

## Mean and Standard Deviation Table for Each Variable

A table is also created displaying the means and standard deviation for each group's reading, writing, and math test scores.

```{r gender_table}
gender_table <- student%>%group_by(gender)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

gender_table
```


```{r ethnicity_table}
#Find the mean and standard deviation of the reading, writing, and math scores for each ethnicity group
ethnic_table <- student%>%group_by(ethnicity)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

ethnic_table
```

```{r education_table}
education_table <- student%>%group_by(parental_education)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

education_table
```


```{r lunch_table}
lunch_table <- student%>%group_by(lunch)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

lunch_table
```

```{r test_prep_table}
test_prep_table <- student%>%group_by(test_prep_course)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

test_prep_table
```

Sometimes we want to see the results of those who fit in multiple groups. Another table is created to show the mean and standard deviation of each gender, further grouped by each gender's race/ethnicity. 

```{r ethnic_gender_table}
ethnic_gender_table <- student%>%group_by(ethnicity, gender)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

ethnic_gender_table
```

The same classification can be done to see the mean and standard deviation of the test results for each combination of test preparation and lunch program groupings. 

```{r test_lunch_table}
test_lunch_table <- student%>%group_by(test_prep_course, lunch)%>%summarize(reading_mean = mean(reading_score, na.rm = TRUE), writing_mean = mean(writing_score, na.rm = TRUE), math_mean = mean(math_score, na.rm = TRUE), reading_sd = sd(reading_score), writing_sd = sd(writing_score), sd_math = sd(math_score))

test_lunch_table
```

From the tables printed, the means seem to differ in each group. The large standard deviations in the tables means the test scores are very spread out, hinting at large variance in the data. The similar standard deviations in each table means we can use t-tests and ANOVA to see if there is a statistical difference in the means. 

## T-Testing

T-testing is effective when comparing the means of two groups. Multiple t-tests can be run to see if there are differences in the testing score means between males and females. To keep the session short, only one t-test is done to demonstrate how a t-test is interpreted. 

In the code below, a t-test is done to see if there is a difference in average reading scores between male and female students. 

```{r t_test}
t.test(reading_score ~ gender, data = student)
```

From the results, we want to see the p-value.  The general idea is that a p-value less than 0.05 means the result is statistically significant, and we can conclude there is a difference in average reading scores between male and female students. From above, we see there is a p-value of 1.49e-08. Since this is less than 0.05, we can conclude there is a difference in the average reading scores of male and female students.

## ANOVA Testing

ANOVA testing is used to check if there is a difference in means among more than two groups. For example, from the boxplot it appears there is a difference in means between the groups in the highest parent education variable. To compare the average reading score in this variable, an ANOVA test is run:

```{r anova}
parent_aov <- aov(reading_score ~ parental_education, data = student)

summary(parent_aov)
```

From the summary, the p-value is 1.49e-08. Since this is less than 0.05, we can conclude there is a difference in the average reading scores between the parent education groups.

## Simple Linear Regression

Linear regression is used to see if there is a relationship between the variables. Normally, linear regression has a dependent variable (y) that can be predicted based on the explanatory variable (x).  Linear regression is simple to implement and on the easier statistical methods to interpret, and will be used in this session.

In this session, simple linear regression is used to see if there is a relationship between two variables. The formula for simple linear regression is $y' = a + bx$, where 'a' is the intercept value (value of y when x = 0) and 'b' is the slope of the line. A positive value for 'b' shows a positive means there's a positive relationship between the variables. In other words, when 'x' increases, so does 'y'. Also, if 'x' decreases, 'y' decreases as well. A negative value for 'b' means there's a negative relationship between the variables. As 'x' increases, 'y' decreases. When 'x' decreases, 'y' will increase.

The simple linear regression I will use will see if there is a relationship between reading score and writing score for each gender.

```{r regression}
ggplot(student, aes(x = reading_score, y = writing_score, color = gender)) + geom_point() + geom_smooth(aes(group = gender),method = "lm") + labs(x = "Reading Score", y = "Writing Score")

male_fit_read_write <- lm(writing_score ~ reading_score, subset = gender == "male", data = student)

summary(male_fit_read_write)

female_fit_read_write <- lm(writing_score ~ reading_score, subset = gender == "female", data = student)

summary(female_fit_read_write)
```

A best fit line, or regression line, is drawn for each gender to show the slope of the formula for each gender. The best fit line also shows the point where the distance between an observed point and the line is minimized.
Linear regression is best used for continuous data, or data with numbers that dare measured (time, distance, weight). Exam scores can only be from 0 to 100 and would be considered discrete data (countable data). This will mean we will have "non-integer" number when looking at the linear regression equation. Linear regression can still be used for exam scores, but the values in the formula will not be whole numbers. 

For example, from the summary it is seen the formula for female students is writing score = 1.21042 + (0.98160) * reading score. So, a female student with a reading score of 50 is predicted to have a writing score of 50.29. A student can't actually score a 50.29 on the exam, it can be interpreted a student with a 50 on the reading score will score around a 50 on the writing score.

From the summary, we can also get the R^2^ value. R^2^ is known as the coefficient of determination, which is used to describe proportion of variance the outcome Y is explained by the regression model. Another way to describe R^2^ is it **measures how well the linear regression model fits the data**. The linear regression model for females has an adjusted R^2^ value = 0.9087, which means 90.87% of the variance can be explained by the model.

The p-value also tells us if the relationship highlighted is statistically significant. The linear regression models above gives a p-value less than 2.2e-16 for both genders, which is significantly less than 0.05. It can be concluded there is a relationship between reading scores and writing scores for both genders. 

After creating a simple linear regression model to see the relationship between the writing score and reading score for each gender, another model is created to see the relationship between reading score and math score for each gender:

```{r regression_2}
ggplot(student, aes(x = reading_score, y = math_score, color = gender))+geom_point() + geom_smooth(aes(group = gender), method = "lm") + labs(x = "Reading Score", y = "Math Score")

male_fit_math_read <- lm(math_score ~ reading_score, subset = gender == "male", data = student)

summary(male_fit_math_read)

female_math_read <- lm(math_score ~ reading_score, subset = gender == "female", data = student)

summary(female_math_read)
```

Looking at the female students linear regression model, the formula is math score = 8.9863 + (0.91247) * reading score. The R^2^ value is 0.8213, so 82.13% of the variance can be explained by the model. The p-value is also less than 2.2e-16 which is less than 0.05, so it can be concluded there is a relationship between math scores and reading scores for female students. 

Another model is created to see the relationship between math score and writing score for each gender:

```{r regression_part_2}
ggplot(student, aes(x = writing_score, y = math_score, color = gender))+geom_point() + geom_smooth(aes(group = gender), method = "lm") + labs(x = "Writing Score", y = "Math Score")

male_fit_math_write <- lm(math_score ~ writing_score, subset = gender == "male", data = student)

summary(male_fit_math_write)

female_math_write <- lm(math_score ~ writing_score, subset = gender == "female", data = student)

summary(female_math_write)
```

Looking at the female students linear regression model, the formula is math score = -5.83959 + (0.95878) * writing score. The R^2^ value is 0.8424, so 84.24% of the variance can be explained by the model. The p-value is also less than 2e-16 which is less than 0.05, so it can be concluded there is a relationship between math scores and reading scores for female students.

## Predict Exam Scores

Since it's been concluded there is a relationship between the test scores for each gender, a prediction model is created to see if a student's test score can be predicted based on another test score and another variable. 


### Factorization

Before setting up the prediction model, the categorical variables must be set up as factors. This way, categorical variables are converted into numeric variables and can be used in the prediction model.

```{r factorization}
student_factor <- as.factor(student)

print(is.factor(student_factor))
```

Now the prediction models can be created in R.

### Exam Score Prediction Models

One prediction model that can be used is the predict function in R. 

```{r predict_test}
#Multiple linear regression model with writing score as the dependent variable and gender, test prep, and gender as the explanatory variables.
test_fit_writing_read <- lm(writing_score ~ gender + reading_score + gender + test_prep_course, data = student)
#Prints the summary of the linear regression model.
summary(test_fit_writing_read)
#Prediction model using the linear regression model. 
predict_test_writing_read <- predict(test_fit_writing_read, interval = "prediction")
#Data frame comparing the prediction model's writing score to the actual writing score. 
actual_pred_test_1 <- data.frame(cbind(prediction = predict_test_writing_read, actual = student$writing_score))
#Lists the first 10 prediction's fit prediction and confidence interval (lwr and upr) and compares the values to the actual writing score. 
head(actual_pred_test_1, 10)
#Views the correlation between the prediction model and the actual model. 
cor(actual_pred_test_1)
```

From the results, the prediction model had relatively close values compared to the actual model's writing score values. The high correlation values (all over 0.95) means the variables in the model have a strong relationship. 

Another prediction that can be used is creating a training data set and a testing data set. A training data set is a subset of examples used to train the model, while the testing data set is a subset used to test the training model. 

```{r test_set}
#Initializes number generator.
set.seed(123)
#New sample created for the training and testing data sets. The data is split with 75% in training and 25% in testing.
sample <- sample(c(TRUE, FALSE), nrow(student), replace = TRUE, prob = c(0.75, 0.25))
train <- student[sample, ]
test <- student[!sample, ]
```

Next, the same linear regression model is set up using the training and testing data sets.

```{r predict_test_2}
#Multiple linear regression model with writing score as the dependent variable and gender, test prep, and gender as the explanatory variables.
lm_practice <- lm(formula = writing_score ~ gender + reading_score + test_prep_course, data = train)
#Prediction model using the linear regression model. 
lm_predict <- predict(lm_practice, newdata = test)
#Data frame comparing the prediction model's writing score to the actual writing score.
actual_pred_1 <- data.frame(cbind(prediction = lm_predict, actual = test$writing_score))
#Lists 10 predictions of the fit prediction model and compares the values to the actual writing score.  
head(actual_pred_1, 10)
#Views the correlation between the prediction model and the actual model. 
cor(actual_pred_1)
```

From the results, the prediction model had relatively close values compared to the actual model's writing score values. The high correlation value (over 0.95) means the variables in the model have a strong relationship.

## Conclusion

In this session, exploratory data analysis techniques, linear regression and prediction modeling were demonstrated using the fictional "Students Performance in Exams" dataset from Kaggle.

First, the dataset was inspected and checked for missing and duplicated values. A frequency table was created to view the sample size of each variable. The data was also checked for outliers, with one outlier chosen to be removed from the dataset. The distribution was viewed and boxplots were created to see the IQR, mean, and outliers in the data.

Next, a mean and standard deviation table were created to compare the testing score means and standard deviations of each group. Since there appears to be differences in the means for each group, a t-test and ANOVA test are performed to confirm. From the t-test and ANOVA test run, it is concluded there are differences in the reading score means for gender and parent highest education. More t-tests and ANOVA tests can be run to see further differences in means, but only these two were completed since this session is used for demonstration purposes.

A linear regression model was completed to demonstrate the relationship between the various testing scores. A linear regression model was done for both male and female students to see the linear regression formula and variance for each gender. 

A prediction model was created in two different ways to see if test scores can be predicted based off of select variables. From both methods, it appears an accurate prediction model was created for predicting writing test scores. Based on this, prediction models can be created for reading and math test scores if desired.

This analysis was meant to demonstrate basic exploratory data analysis techniques as well as linear regression and prediction modeling. For a real world data set, I would delve deeper into the data using more complex statistical methods. Such statistical methods can include logistic regression, random forest modeling, and clustering.

Thank you to the reader for viewing my work. 

## END